---
title: "Century Health — Real World Evidence: Lupus Patient Analysis"
author: "Data Analyst Take-Home"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# 0. Setup & Data Loading

```{r load-libraries}
library(tidyverse)
library(lubridate)
library(readxl)
library(scales)
library(patchwork)
library(broom)
library(testthat)
library(knitr)

# Load modular helper functions
source("R/helpers.R")

cat("All libraries and helpers loaded.\n")
```

```{r load-data}
raw <- load_all_datasets("data")

cat(sprintf("Patients:    %d x %d\n", nrow(raw$patients),    ncol(raw$patients)))
cat(sprintf("Encounters:  %d x %d\n", nrow(raw$encounters),  ncol(raw$encounters)))
cat(sprintf("Symptoms:    %d x %d\n", nrow(raw$symptoms),    ncol(raw$symptoms)))
cat(sprintf("Medications: %d x %d\n", nrow(raw$medications), ncol(raw$medications)))
cat(sprintf("Conditions:  %d x %d\n", nrow(raw$conditions),  ncol(raw$conditions)))
```

---

# 1. Data Assessment

## 1.1 Entity-Relationship Diagram

The five datasets are connected through **patient UUIDs** and **encounter UUIDs**:

```
                          ┌──────────────┐
                          │   PATIENTS   │
                          │ (PATIENT_ID) │
                          └──────┬───────┘
                                 │  1
              ┌──────────────────┼──────────────────┐
              │                  │                  │
              ▼ *                ▼ *                ▼ *
     ┌────────────────┐  ┌──────────────┐  ┌──────────────┐
     │  ENCOUNTERS    │  │  SYMPTOMS    │  │  MEDICATIONS │
     │  (Id)          │  │  (PATIENT)   │  │  (PATIENT,   │
     │  FK: PATIENT   │  │              │  │   ENCOUNTER) │
     └───────┬────────┘  └──────────────┘  └──────┬───────┘
             │ 1                                  │
             │            ┌──────────────┐        │
             └───────────►│  CONDITIONS  │◄───────┘
                     *    │  (PATIENT,   │    *
                          │   ENCOUNTER) │
                          └──────────────┘
```

| From | To | Join Key |
|------|------|----------|
| Patients → Encounters | `PATIENT_ID` = `PATIENT` |
| Patients → Symptoms | `PATIENT_ID` = `PATIENT` |
| Patients → Medications | `PATIENT_ID` = `PATIENT` |
| Patients → Conditions | `PATIENT_ID` = `PATIENT` |
| Encounters → Medications | `Id` = `ENCOUNTER` |
| Encounters → Conditions | `Id` = `ENCOUNTER` |

> **Note on file formats:** The raw encounters data is provided as Parquet. Because the R `arrow` package is not always available, a pre-converted `encounters.csv` is included in the submission. See `R/helpers.R` for regeneration instructions.

## 1.2 Key Data Quality Issues Identified

| # | Issue | Impact | Resolution |
|---|-------|--------|------------|
| 1 | Mixed-case UUID keys across files | Joins fail silently | Normalise all IDs to lowercase |
| 2 | Same medication in 3+ casing variants | Inflated distinct-med counts | Standardise to uppercase |
| 3 | GENDER column entirely NA in symptoms | Missing demographic info | Backfill from patients table |
| 4 | Date columns stored as strings | Can't do time-based analysis | Parse to Date / POSIXct |
| 5 | SYMPTOMS is a packed `key:value` string | Not analysable directly | Parse into 4 numeric columns |
| 6 | Symptoms table has duplicate rows per patient | Overstates sample size, mixes pathologies | Filter to Lupus pathology, deduplicate to 1 row/patient |
| 7 | File format heterogeneity (CSV, Parquet, XLSX) | Loading complexity | Handled in `load_all_datasets()` |

## 1.3 Two Cohort Definitions

The data supports two natural cohort definitions for Lupus patients. We label them clearly since they have different sizes and are used in different parts of this analysis:

| Cohort | Definition | Used in |
|--------|-----------|---------|
| **Diagnosed cohort** | Patients with >= 1 Lupus erythematosus record in `conditions` | Sections 3.1–3.3 |
| **Symptom-observed subset** | Diagnosed patients who also have a Lupus row in `symptoms` | Sections 3.4, 4, 5 |

The **symptom-observed subset** is a strict subset of the diagnosed cohort. Exact cohort sizes are computed during cleaning and reported in Section 2.

---

# 2. Data Cleaning

```{r clean-data}
# Step 1: Normalise IDs to lowercase
datasets <- normalise_ids(raw)

# Step 2: Standardise medication names
datasets$medications <- standardise_medication_names(datasets$medications)
cat(sprintf("Medication descriptions: %d unique (after normalisation)\n",
            n_distinct(datasets$medications$DESCRIPTION)))

# Step 3: Parse dates (including DEATHDATE and conditions$STOP)
datasets <- parse_dates(datasets)

# Step 4: Parse symptom scores
datasets$symptoms <- parse_symptom_scores(datasets$symptoms)

# Step 5: Filter to Lupus-only rows and deduplicate to one row per patient
cat(sprintf("Symptoms before Lupus filter: %d rows, %d unique patients\n",
            nrow(datasets$symptoms), n_distinct(datasets$symptoms$PATIENT)))
datasets$symptoms <- filter_lupus_symptoms(datasets$symptoms)
cat(sprintf("Symptoms after  Lupus filter: %d rows, %d unique patients\n",
            nrow(datasets$symptoms), n_distinct(datasets$symptoms$PATIENT)))

# Step 6: Fill missing GENDER in symptoms from patients table
datasets$symptoms <- fill_symptom_gender(datasets$symptoms, datasets$patients)
cat(sprintf("Symptom GENDER nulls remaining: %d\n",
            sum(is.na(datasets$symptoms$GENDER))))

# Show parsed symptom scores
datasets$symptoms %>%
  select(PATIENT, PATHOLOGY, Rash, Joint_Pain, Fatigue, Fever) %>%
  head(5) %>%
  kable(caption = "Parsed Lupus symptom scores (first 5 rows)")
```

## 2.1 Data Integrity Unit Tests

```{r unit-tests, results='asis'}
run_integrity_tests(datasets)
cat("\n**All data integrity tests passed.**\n")
```

```{r extract-for-convenience}
# Extract cleaned tables for downstream analysis
patients    <- datasets$patients
encounters  <- datasets$encounters
symptoms    <- datasets$symptoms
medications <- datasets$medications
conditions  <- datasets$conditions

# ── Define the two cohorts ──
# Diagnosed cohort: all patients with a Lupus condition record
lupus_patient_ids <- unique(conditions %>%
  filter(str_to_upper(DESCRIPTION) == "LUPUS ERYTHEMATOSUS") %>%
  pull(PATIENT))

lupus_patients <- patients %>% filter(PATIENT_ID %in% lupus_patient_ids)

# Symptom-observed subset: diagnosed patients who also have symptom data
symptom_patient_ids <- unique(symptoms$PATIENT)

n_total     <- n_distinct(patients$PATIENT_ID)
n_diagnosed <- n_distinct(lupus_patients$PATIENT_ID)
n_symptom   <- n_distinct(symptom_patient_ids)

# Dynamic cohort-size table (referenced from Section 1.3)
tibble(
  Cohort     = c("All patients in raw dataset",
                 "Diagnosed Lupus cohort",
                 "Symptom-observed Lupus subset"),
  Definition = c("Distinct patients in `patients.csv`",
                 "Patients with >= 1 Lupus erythematosus record in `conditions`",
                 "Diagnosed patients who also have a Lupus row in `symptoms`"),
  Size       = c(comma(n_total), comma(n_diagnosed), comma(n_symptom))
) %>% kable(caption = "Cohort definitions and sizes")
```

---

# 3. Descriptive Analysis

## 3.1 How Many Distinct Patients Are in the Dataset?

```{r distinct-patients}
cat(sprintf("Total distinct patients in the raw dataset: %s\n", comma(n_total)))
cat(sprintf("  Diagnosed Lupus cohort:                   %s\n", comma(n_diagnosed)))
cat(sprintf("  Symptom-observed Lupus subset:            %s\n", comma(n_symptom)))
```

The raw `patients.csv` file contains `r comma(n_total)` distinct patients. Of these, `r comma(n_diagnosed)` have at least one Lupus erythematosus diagnosis in the conditions table, and `r comma(n_symptom)` also have Lupus symptom records. The remaining descriptive statistics in this section use the **diagnosed cohort** unless noted otherwise.

## 3.2 Distinct Medications Over Time (Diagnosed Cohort)

This chart counts only **Lupus-directed medications** — prescriptions where `REASONDESCRIPTION` is "Lupus erythematosus." This deliberately excludes supportive-care medications (e.g. Vitamin B12, ferrous sulfate) that are prescribed for comorbid anemia rather than Lupus itself.

```{r meds-over-time, fig.height=4}
# Lupus-directed medications: prescribed to diagnosed patients AND
# linked to Lupus via REASONDESCRIPTION (excludes supportive-care meds
# like Vitamin B12 and ferrous sulfate for anemia)
lupus_meds <- medications %>%
  filter(PATIENT %in% lupus_patient_ids,
         str_to_upper(REASONDESCRIPTION) == "LUPUS ERYTHEMATOSUS")

meds_over_time <- lupus_meds %>%
  mutate(YEAR_MONTH = floor_date(START, "month")) %>%
  group_by(YEAR_MONTH) %>%
  summarise(distinct_meds = n_distinct(DESCRIPTION), .groups = "drop")

ggplot(meds_over_time, aes(x = YEAR_MONTH, y = distinct_meds)) +
  geom_line(linewidth = 0.7, colour = "#2c7fb8") +
  geom_point(size = 1.5, colour = "#2c7fb8") +
  scale_y_continuous(breaks = scales::pretty_breaks()) +
  labs(
    title = "Distinct Medications Prescribed per Month (Diagnosed Lupus Cohort)",
    x     = "Date",
    y     = "Distinct Medications"
  ) +
  theme_minimal()
```

## 3.3 Diagnosed Cohort Distribution by Race and Gender

```{r pie-charts, fig.height=5}
# Race pie — diagnosed Lupus cohort
race_data <- lupus_patients %>%
  count(RACE) %>%
  mutate(pct = n / sum(n),
         label = sprintf("%s\n%.1f%%", RACE, pct * 100))

p_race <- ggplot(race_data, aes(x = "", y = n, fill = RACE)) +
  geom_col(width = 1, colour = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5), size = 3) +
  labs(title = "Diagnosed Lupus Patients by Race", fill = "Race") +
  theme_void()

# Gender pie — diagnosed Lupus cohort
gender_data <- lupus_patients %>%
  count(GENDER) %>%
  mutate(pct = n / sum(n),
         label = sprintf("%s\n%.1f%%", GENDER, pct * 100))

p_gender <- ggplot(gender_data, aes(x = "", y = n, fill = GENDER)) +
  geom_col(width = 1, colour = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5), size = 3.5) +
  labs(title = "Diagnosed Lupus Patients by Gender", fill = "Gender") +
  theme_void()

p_race + p_gender
```

## 3.4 Patients with All Four Symptom Categories >= 30 (Symptom-Observed Subset)

```{r high-symptoms}
# symptoms is one row per patient in the symptom-observed subset
all_high <- symptoms %>%
  filter(Rash >= 30, Joint_Pain >= 30, Fatigue >= 30, Fever >= 30)

pct_high <- nrow(all_high) / nrow(symptoms) * 100

cat(sprintf(
  "Symptom-observed patients with ALL four scores >= 30: %d / %d (%.1f%%)\n",
  nrow(all_high), nrow(symptoms), pct_high
))
cat(sprintf("(Note: this uses the symptom-observed subset of %s patients, not the full diagnosed cohort.)\n",
            comma(n_symptom)))
```

---

# 4. Cohort Analysis — Treatments, Demographics & Outcomes

> **Cohort scope:** This section uses the **symptom-observed subset** (`r comma(n_symptom)` patients) since symptom scores are required for outcome measurement.

## Analysis Plan

We examine how **medication type**, **ethnicity**, and **age group** relate to symptom severity.

**Cohort definitions:**

| Dimension | Groups |
|-----------|--------|
| Ethnicity | Hispanic, Non-Hispanic |
| Age Group | Young (<30), Middle (30–54), Older (55+) |
| Medication | Determined by each patient's most-prescribed therapy (see below) |

**Outcome measure:** Mean composite symptom score (mean of Rash, Joint Pain, Fatigue, Fever).

**Approach:**

1. Classify each medication into a therapeutic group.
2. Identify each patient's primary therapy (most frequently prescribed; ties broken alphabetically by therapy name).
3. Report which therapies actually appear as primary in this subset.
4. Assign age groups and compute composite symptom scores.
5. Compare outcomes across therapy × ethnicity × age group.

```{r build-cohort}
# Classify therapies (all Lupus medications)
lupus_meds <- lupus_meds %>%
  mutate(THERAPY = classify_therapy(DESCRIPTION))

cat("Therapy distribution (all Lupus cohort prescriptions):\n")
lupus_meds %>% count(THERAPY, sort = TRUE) %>% kable()
```

```{r build-analysis-cohort}
# Restrict medications to the symptom-observed subset
sym_meds <- lupus_meds %>% filter(PATIENT %in% symptom_patient_ids)

# Each patient's primary therapy — deterministic tie-breaking:
#   1. Most prescriptions wins
#   2. Ties broken alphabetically by therapy name
patient_therapy <- sym_meds %>%
  count(PATIENT, THERAPY, name = "n_prescriptions") %>%
  arrange(PATIENT, desc(n_prescriptions), THERAPY) %>%
  group_by(PATIENT) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  select(PATIENT, THERAPY)

# Report which therapies actually survived as primary
cat("Primary therapy distribution (symptom-observed subset):\n")
patient_therapy %>% count(THERAPY, sort = TRUE) %>% kable()
```

```{r define-cohort-main}
# Add composite score and age group to the deduplicated symptoms table
symptoms <- symptoms %>%
  add_composite_score() %>%
  mutate(AGE_GROUP = assign_age_group(AGE_BEGIN))

# Join symptoms with primary therapy
cohort <- symptoms %>%
  inner_join(patient_therapy, by = "PATIENT")

# Identify which therapies have enough patients for meaningful comparison
therapy_counts <- cohort %>% count(THERAPY, sort = TRUE)
included_therapies <- therapy_counts %>% filter(n >= 10) %>% pull(THERAPY)

cohort_main <- cohort %>% filter(THERAPY %in% included_therapies)

cat(sprintf("\nAnalysis cohort: %d patients (one row each)\n", nrow(cohort_main)))
cat("Therapy groups included (n >= 10):\n")
cohort_main %>% count(THERAPY, sort = TRUE) %>% kable()
```

## 4.1 Symptom Severity by Therapy and Age Group

```{r therapy-age-bar, fig.height=5}
summary_age <- cohort_main %>%
  group_by(THERAPY, AGE_GROUP) %>%
  summarise(mean_score = mean(COMPOSITE_SCORE), .groups = "drop")

ggplot(summary_age, aes(x = THERAPY, y = mean_score, fill = AGE_GROUP)) +
  geom_col(position = "dodge", colour = "black", linewidth = 0.3) +
  labs(
    title = "Mean Composite Symptom Score by Therapy and Age Group",
    x     = "Therapy",
    y     = "Mean Composite Score",
    fill  = "Age Group"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 10, hjust = 1))
```

## 4.2 Symptom Severity by Therapy and Ethnicity

```{r therapy-ethnicity-bar, fig.height=5}
summary_eth <- cohort_main %>%
  group_by(THERAPY, ETHNICITY) %>%
  summarise(mean_score = mean(COMPOSITE_SCORE), .groups = "drop")

ggplot(summary_eth, aes(x = THERAPY, y = mean_score, fill = ETHNICITY)) +
  geom_col(position = "dodge", colour = "black", linewidth = 0.3) +
  labs(
    title = "Mean Composite Symptom Score by Therapy and Ethnicity",
    x     = "Therapy",
    y     = "Mean Composite Score",
    fill  = "Ethnicity"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 10, hjust = 1))
```

## 4.3 Heatmap: Therapy × Age Group × Ethnicity

```{r heatmap, fig.height=4.5}
cohort_main <- cohort_main %>%
  mutate(COHORT_LABEL = paste(AGE_GROUP, "/", ETHNICITY))

heat_data <- cohort_main %>%
  group_by(THERAPY, COHORT_LABEL) %>%
  summarise(mean_score = mean(COMPOSITE_SCORE), .groups = "drop")

ggplot(heat_data, aes(x = COHORT_LABEL, y = THERAPY, fill = mean_score)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.1f", mean_score)), size = 3) +
  scale_fill_gradient(low = "#ffffcc", high = "#d7191c") +
  labs(
    title = "Mean Composite Symptom Score: Therapy x Cohort",
    x     = "Cohort (Age Group / Ethnicity)",
    y     = "Therapy",
    fill  = "Mean Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## 4.4 Individual Symptom Distributions by Therapy

```{r boxplots, fig.height=8}
symptom_long <- cohort_main %>%
  pivot_longer(cols = c(Rash, Joint_Pain, Fatigue, Fever),
               names_to = "Symptom", values_to = "Score")

ggplot(symptom_long, aes(x = THERAPY, y = Score, fill = THERAPY)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~ Symptom, scales = "free_y", ncol = 2) +
  labs(
    title = "Symptom Score Distributions by Therapy",
    x     = NULL,
    y     = "Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom")
```

## 4.5 Discussion

```{r cohort-summary-stats}
# Summarise actual values so the narrative is grounded in outputs
overall_means <- cohort_main %>%
  group_by(THERAPY) %>%
  summarise(
    n = n(),
    mean_composite = round(mean(COMPOSITE_SCORE), 1),
    sd_composite   = round(sd(COMPOSITE_SCORE), 1),
    .groups = "drop"
  )

kable(overall_means,
      col.names = c("Therapy", "N", "Mean Composite", "SD"),
      caption   = "Summary statistics by therapy")

# Check whether age-group or ethnicity differences are large
age_spread <- summary_age %>%
  group_by(THERAPY) %>%
  summarise(range = round(max(mean_score) - min(mean_score), 1), .groups = "drop")

eth_spread <- summary_eth %>%
  group_by(THERAPY) %>%
  summarise(range = round(max(mean_score) - min(mean_score), 1), .groups = "drop")

cat("\nMax within-therapy difference across age groups:\n")
print(age_spread)
cat("\nMax within-therapy difference across ethnicities:\n")
print(eth_spread)
```

**Observations based on the summaries above:**

- The table above shows the actual mean composite scores and standard deviations for each therapy group that survived primary-therapy assignment. If the means are similar and the SDs overlap, this indicates no observable treatment differentiation.
- The within-therapy spread across age groups and ethnicities is reported above. Small ranges (a few points) indicate that demographic subgroups do not meaningfully differ within a therapy.
- The box plots show the full distributions. If the medians and interquartile ranges overlap substantially, this further supports the absence of a treatment-specific signal.
- These patterns are consistent with Synthea's simulation approach, which generates symptom scores independently of treatment assignment.

---

# 5. Statistical Analysis

## Goal

The assignment asks for a model of whether a symptom outcome decreases after a chosen time period following medication. However, the symptom data is **cross-sectional** — each patient has a single symptom observation, not a before/after pair. There is no longitudinal follow-up in the symptoms table.

**Limitation and fallback:** Because the data does not support a genuine pre/post comparison, we instead model a cross-sectional outcome: whether Fatigue is **below the cohort median** (a proxy for relatively low severity). This is a weaker analysis than the assignment requests, but it is the strongest model the data supports without fabricating a temporal structure.

**Predictors:** therapy type, age, ethnicity, race, gender, total prescription count.

**Train/test split:** 75/25 random split at the patient level (one row per patient after deduplication, so no leakage risk).

> **Cohort scope:** Symptom-observed subset only.

```{r model-prep}
# Count total prescriptions per patient (within symptom-observed subset)
rx_counts <- sym_meds %>%
  count(PATIENT, name = "N_PRESCRIPTIONS")

# Build modelling dataset
model_df <- cohort_main %>%
  left_join(rx_counts, by = "PATIENT") %>%
  filter(!is.na(GENDER), !is.na(RACE), !is.na(ETHNICITY)) %>%
  mutate(
    LOW_FATIGUE = as.integer(Fatigue < median(Fatigue)),
    THERAPY     = droplevels(factor(THERAPY)),
    GENDER      = droplevels(factor(GENDER)),
    # Collapse sparse race levels (< 20 patients) into "other"
    RACE        = fct_lump_min(factor(RACE), min = 20, other_level = "other"),
    RACE        = relevel(RACE, ref = "white"),
    ETHNICITY   = droplevels(factor(ETHNICITY))
  )

cat("Race levels after collapsing sparse categories:\n")
table(model_df$RACE) %>% print()

# Drop factor columns with fewer than 2 levels (can't be used in regression)
single_level_cols <- names(model_df)[sapply(model_df, function(x) {
  is.factor(x) && nlevels(x) < 2
})]
if (length(single_level_cols) > 0) {
  cat(sprintf("Dropping single-level factors: %s\n",
              paste(single_level_cols, collapse = ", ")))
  model_df <- model_df %>% select(-all_of(single_level_cols))
}

cat(sprintf("Modelling dataset: %d patients\n", nrow(model_df)))
cat(sprintf("Fatigue median threshold: %.0f\n", median(model_df$Fatigue)))
cat("Outcome distribution (LOW_FATIGUE):\n")
table(model_df$LOW_FATIGUE) %>% print()
```

```{r fit-model}
# Build formula dynamically (skip any dropped single-level factors)
candidate_predictors <- c("AGE_BEGIN", "GENDER", "RACE", "ETHNICITY",
                          "THERAPY", "N_PRESCRIPTIONS")
available_predictors <- intersect(candidate_predictors, names(model_df))
model_formula <- as.formula(
  paste("LOW_FATIGUE ~", paste(available_predictors, collapse = " + "))
)
cat("Model formula:", deparse(model_formula), "\n\n")

# Fit logistic regression on full dataset for coefficient inspection
logit_fit <- glm(model_formula, data = model_df, family = binomial(link = "logit"))

summary(logit_fit)
```

## 5.1 Odds Ratios and Confidence Intervals

```{r odds-ratios}
or_table <- tidy(logit_fit, conf.int = TRUE, exponentiate = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  rename(
    `Odds Ratio`     = estimate,
    `CI Lower (95%)` = conf.low,
    `CI Upper (95%)` = conf.high,
    `p-value`        = p.value
  )

kable(or_table, digits = 3, caption = "Logistic Regression — Odds Ratios")
```

## 5.2 Model Evaluation

```{r model-eval, fig.height=5}
# Random train/test split at patient level (one row per patient)
set.seed(42)
train_idx <- sample(seq_len(nrow(model_df)), size = floor(0.75 * nrow(model_df)))
train_df  <- model_df[train_idx, ]
test_df   <- model_df[-train_idx, ]

cat(sprintf("Train: %d patients, Test: %d patients\n",
            nrow(train_df), nrow(test_df)))

# Refit on training data
logit_train <- glm(model_formula, data = train_df, family = binomial(link = "logit"))

# Predict on test data
test_df$pred_prob  <- predict(logit_train, newdata = test_df, type = "response")
test_df$pred_class <- as.integer(test_df$pred_prob >= 0.5)

# Confusion matrix
cat("\nConfusion Matrix:\n")
table(Predicted = test_df$pred_class, Actual = test_df$LOW_FATIGUE) %>% print()

# Accuracy
acc <- mean(test_df$pred_class == test_df$LOW_FATIGUE)
cat(sprintf("\nAccuracy: %.1f%%\n", acc * 100))

# ROC Curve
roc_data <- tibble(
  truth = test_df$LOW_FATIGUE,
  prob  = test_df$pred_prob
) %>%
  arrange(desc(prob)) %>%
  mutate(
    tpr = cumsum(truth == 1) / sum(truth == 1),
    fpr = cumsum(truth == 0) / sum(truth == 0)
  )

auc_val <- abs(sum(diff(roc_data$fpr) *
  (head(roc_data$tpr, -1) + tail(roc_data$tpr, -1)) / 2))

ggplot(roc_data, aes(x = fpr, y = tpr)) +
  geom_line(colour = "#2c7fb8", linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 0.6, y = 0.3,
           label = sprintf("AUC = %.3f", auc_val), size = 4.5) +
  labs(
    title = "ROC Curve — Low Fatigue Classification",
    x     = "False Positive Rate",
    y     = "True Positive Rate"
  ) +
  theme_minimal()
```

## 5.3 Model Interpretation

**How to read the output:**

- **Coefficients (log-odds):** Each coefficient represents the change in log-odds of having low Fatigue for a one-unit increase in that predictor, holding all others constant.

- **Odds Ratios:** OR > 1 means the predictor increases the odds of low Fatigue relative to the reference category; OR < 1 decreases it.

- **p-values:** Values below 0.05 indicate statistical significance at the 5% level.

- **Confidence Intervals:** If the 95% CI for an odds ratio crosses 1.0, the association is not statistically significant.

```{r interpret-results, results='asis'}
# Dynamically report findings based on actual model output
sig_terms <- or_table %>%
  filter(term != "(Intercept)", `p-value` < 0.05)
n_sig <- nrow(sig_terms)

cat("**Key findings from this model:**\n\n")

if (n_sig == 0) {
  cat("- No predictor reaches statistical significance at the 5% level.\n")
  cat("- This means therapy type, demographics, and prescription volume are ",
      "not meaningfully associated with Fatigue level in this cohort.\n")
} else {
  cat(sprintf("- %d predictor(s) reached significance at p < 0.05:\n", n_sig))
  for (i in seq_len(nrow(sig_terms))) {
    cat(sprintf("  - **%s**: OR = %.2f (95%% CI: %.2f\u2013%.2f, p = %.3f)\n",
                sig_terms$term[i], sig_terms$`Odds Ratio`[i],
                sig_terms$`CI Lower (95%)`[i], sig_terms$`CI Upper (95%)`[i],
                sig_terms$`p-value`[i]))
  }
}

cat(sprintf("\n- **AUC = %.3f** \u2014 ", auc_val))
if (auc_val < 0.55) {
  cat("near 0.5, indicating the model has no meaningful discriminatory power.\n")
} else if (auc_val < 0.7) {
  cat("indicating weak discriminatory power.\n")
} else {
  cat("indicating moderate to good discriminatory power.\n")
}

cat(sprintf("- **Accuracy = %.1f%%** on the held-out test set.\n\n", acc * 100))

cat("**Limitations:**\n\n")
cat("- **Cross-sectional design:** The outcome (Fatigue below vs. above median) ",
    "is not a measure of change over time. The assignment asks for a model of ",
    "whether an outcome decreases after medication; however, the symptom data ",
    "provides only a single observation per patient with no pre-treatment baseline. ",
    "A genuine before/after analysis would require longitudinal symptom records.\n")
cat("- **Observational data:** Patients were not randomly assigned to therapies, ",
    "so any associations may be confounded by indication.\n")
cat("- **Simulated data:** All patterns (or lack thereof) reflect Synthea's ",
    "data-generating process rather than real clinical relationships.\n")
```
