---
title: "Century Health — Real World Evidence: Lupus Patient Analysis"
author: "Data Analyst Take-Home"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# 0. Setup & Data Loading

```{r load-libraries}
library(tidyverse)
library(readxl)
library(scales)
library(patchwork)
library(broom)
library(testthat)
library(knitr)

# Load modular helper functions
source("R/helpers.R")

cat("All libraries and helpers loaded.\n")
```

```{r load-data}
raw <- load_all_datasets("data")

cat(sprintf("Patients:    %d x %d\n", nrow(raw$patients),    ncol(raw$patients)))
cat(sprintf("Encounters:  %d x %d\n", nrow(raw$encounters),  ncol(raw$encounters)))
cat(sprintf("Symptoms:    %d x %d\n", nrow(raw$symptoms),    ncol(raw$symptoms)))
cat(sprintf("Medications: %d x %d\n", nrow(raw$medications), ncol(raw$medications)))
cat(sprintf("Conditions:  %d x %d\n", nrow(raw$conditions),  ncol(raw$conditions)))
```

---

# 1. Data Assessment

## 1.1 Entity-Relationship Diagram

The five datasets are connected through **patient UUIDs** and **encounter UUIDs**:

```
                          ┌──────────────┐
                          │   PATIENTS   │
                          │ (PATIENT_ID) │
                          └──────┬───────┘
                                 │  1
              ┌──────────────────┼──────────────────┐
              │                  │                  │
              ▼ *                ▼ *                ▼ *
     ┌────────────────┐  ┌──────────────┐  ┌──────────────┐
     │  ENCOUNTERS    │  │  SYMPTOMS    │  │  MEDICATIONS │
     │  (Id)          │  │  (PATIENT)   │  │  (PATIENT,   │
     │  FK: PATIENT   │  │              │  │   ENCOUNTER) │
     └───────┬────────┘  └──────────────┘  └──────┬───────┘
             │ 1                                  │
             │            ┌──────────────┐        │
             └───────────►│  CONDITIONS  │◄───────┘
                     *    │  (PATIENT,   │    *
                          │   ENCOUNTER) │
                          └──────────────┘
```

| From | To | Join Key |
|------|------|----------|
| Patients → Encounters | `PATIENT_ID` = `PATIENT` |
| Patients → Symptoms | `PATIENT_ID` = `PATIENT` |
| Patients → Medications | `PATIENT_ID` = `PATIENT` |
| Patients → Conditions | `PATIENT_ID` = `PATIENT` |
| Encounters → Medications | `Id` = `ENCOUNTER` |
| Encounters → Conditions | `Id` = `ENCOUNTER` |

> **Note:** File formats vary across datasets — CSV, Parquet (converted to CSV for R), and Excel — all handled by the `load_all_datasets()` helper.

---

# 2. Data Cleaning

Issues identified and fixed:

1. **Mixed-case UUID keys** — Patient and Encounter IDs use different casing across files (lowercase in patients/symptoms, UPPERCASE in encounters/conditions). Normalised to lowercase.
2. **Inconsistent medication descriptions** — Same drug appears as `predniSONE 20 MG Oral Tablet`, `PREDNISONE 20 MG ORAL TABLET`, and `prednisone 20 mg oral tablet`. Standardised to uppercase.
3. **Missing GENDER in symptoms** — The GENDER column is entirely NA; backfilled from the patients table.
4. **Date columns stored as strings** — Parsed to proper date/datetime types.
5. **SYMPTOMS column is a packed string** — Parsed `"Rash:34;Joint Pain:39;Fatigue:9;Fever:12"` into four numeric columns.
6. **File format heterogeneity** — CSV, Parquet, and Excel files all handled during loading.

```{r clean-data}
# Step 1: Normalise IDs to lowercase
datasets <- normalise_ids(raw)

# Step 2: Standardise medication names
datasets$medications <- standardise_medication_names(datasets$medications)
cat(sprintf("Medication descriptions: %d unique (after normalisation)\n",
            n_distinct(datasets$medications$DESCRIPTION)))

# Step 3: Parse dates
datasets <- parse_dates(datasets)

# Step 4: Parse symptom scores
datasets$symptoms <- parse_symptom_scores(datasets$symptoms)
datasets$symptoms %>%
  select(PATIENT, Rash, Joint_Pain, Fatigue, Fever) %>%
  head(5) %>%
  kable(caption = "Parsed symptom scores (first 5 rows)")

# Step 5: Fill missing GENDER in symptoms
datasets$symptoms <- fill_symptom_gender(datasets$symptoms, datasets$patients)
cat(sprintf("Symptom GENDER nulls remaining: %d\n",
            sum(is.na(datasets$symptoms$GENDER))))
```

## 2.1 Data Integrity Unit Tests

```{r unit-tests, results='asis'}
run_integrity_tests(datasets)
cat("\n**All data integrity tests passed.**\n")
```

```{r extract-for-convenience}
# Extract cleaned tables for convenience in downstream analysis
patients    <- datasets$patients
encounters  <- datasets$encounters
symptoms    <- datasets$symptoms
medications <- datasets$medications
conditions  <- datasets$conditions
```

---

# 3. Descriptive Analysis

## 3.1 Distinct Patients

```{r distinct-patients}
n_patients <- n_distinct(patients$PATIENT_ID)
cat(sprintf("Distinct patients in the dataset: %s\n", comma(n_patients)))
```

## 3.2 Distinct Medications Over Time

```{r meds-over-time, fig.height=4}
meds_over_time <- medications %>%
  mutate(YEAR_MONTH = floor_date(START, "month")) %>%
  group_by(YEAR_MONTH) %>%
  summarise(distinct_meds = n_distinct(DESCRIPTION), .groups = "drop")

ggplot(meds_over_time, aes(x = YEAR_MONTH, y = distinct_meds)) +
  geom_line(linewidth = 0.7, colour = "#2c7fb8") +
  geom_point(size = 1.5, colour = "#2c7fb8") +
  scale_y_continuous(breaks = scales::pretty_breaks()) +
  labs(
    title = "Distinct Medications Prescribed per Month",
    x     = "Date",
    y     = "Distinct Medications"
  ) +
  theme_minimal()
```

## 3.3 Patient Distribution by Race and Gender

```{r pie-charts, fig.height=5}
# Race pie
race_data <- patients %>%
  count(RACE) %>%
  mutate(pct = n / sum(n),
         label = sprintf("%s\n%.1f%%", RACE, pct * 100))

p_race <- ggplot(race_data, aes(x = "", y = n, fill = RACE)) +
  geom_col(width = 1, colour = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5), size = 3) +
  labs(title = "Patients by Race", fill = "Race") +
  theme_void()

# Gender pie
gender_data <- patients %>%
  count(GENDER) %>%
  mutate(pct = n / sum(n),
         label = sprintf("%s\n%.1f%%", GENDER, pct * 100))

p_gender <- ggplot(gender_data, aes(x = "", y = n, fill = GENDER)) +
  geom_col(width = 1, colour = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5), size = 3.5) +
  labs(title = "Patients by Gender", fill = "Gender") +
  theme_void()

p_race + p_gender
```

## 3.4 Patients with All Four Symptom Categories >= 30

```{r high-symptoms}
all_high <- symptoms %>%
  filter(Rash >= 30, Joint_Pain >= 30, Fatigue >= 30, Fever >= 30)

pct_high <- nrow(all_high) / nrow(symptoms) * 100

cat(sprintf(
  "Patients with ALL four symptom scores >= 30: %d / %d (%.1f%%)\n",
  nrow(all_high), nrow(symptoms), pct_high
))
```

---

# 4. Cohort Analysis — Treatments, Demographics & Outcomes

## Analysis Plan

We examine how **medication type**, **ethnicity**, and **age group** relate to symptom severity.

**Cohort definitions:**

| Dimension | Groups |
|-----------|--------|
| Medication | Naproxen (NSAID), Prednisone (Corticosteroid), Cyclosporine (Immunosuppressant) |
| Ethnicity | Hispanic, Non-Hispanic |
| Age Group | Young (<30), Middle (30–54), Older (55+) |

**Outcome measure:** Mean composite symptom score (average of Rash, Joint Pain, Fatigue, Fever).

**Approach:**

1. Classify each medication into a therapeutic group.
2. Identify each patient's primary therapy (most frequently prescribed).
3. Assign age groups and compute composite symptom scores.
4. Compare outcomes across therapy x ethnicity x age group.

```{r build-cohort}
# Classify therapies
medications <- medications %>%
  mutate(THERAPY = classify_therapy(DESCRIPTION))

cat("Therapy distribution:\n")
medications %>% count(THERAPY, sort = TRUE) %>% kable()
```

```{r build-analysis-cohort}
# Each patient's primary therapy (most frequently prescribed)
patient_therapy <- medications %>%
  count(PATIENT, THERAPY, name = "n_prescriptions") %>%
  arrange(desc(n_prescriptions)) %>%
  distinct(PATIENT, .keep_all = TRUE) %>%
  select(PATIENT, THERAPY)

# Add composite score and age group
symptoms <- symptoms %>%
  add_composite_score() %>%
  mutate(AGE_GROUP = assign_age_group(AGE_BEGIN))

# Join symptoms with primary therapy
cohort <- symptoms %>%
  inner_join(patient_therapy, by = "PATIENT")

# Focus on the three main Lupus therapies
main_therapies <- c("Naproxen (NSAID)", "Prednisone (Corticosteroid)",
                    "Cyclosporine (Immunosuppressant)")
cohort_main <- cohort %>% filter(THERAPY %in% main_therapies)

cat(sprintf("Cohort: %d rows, %d unique patients\n",
            nrow(cohort_main), n_distinct(cohort_main$PATIENT)))
```

## 4.1 Symptom Severity by Therapy and Age Group

```{r therapy-age-bar, fig.height=5}
summary_age <- cohort_main %>%
  group_by(THERAPY, AGE_GROUP) %>%
  summarise(mean_score = mean(COMPOSITE_SCORE), .groups = "drop")

ggplot(summary_age, aes(x = THERAPY, y = mean_score, fill = AGE_GROUP)) +
  geom_col(position = "dodge", colour = "black", linewidth = 0.3) +
  labs(
    title = "Mean Composite Symptom Score by Therapy and Age Group",
    x     = "Therapy",
    y     = "Mean Composite Score",
    fill  = "Age Group"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 10, hjust = 1))
```

## 4.2 Symptom Severity by Therapy and Ethnicity

```{r therapy-ethnicity-bar, fig.height=5}
summary_eth <- cohort_main %>%
  group_by(THERAPY, ETHNICITY) %>%
  summarise(mean_score = mean(COMPOSITE_SCORE), .groups = "drop")

ggplot(summary_eth, aes(x = THERAPY, y = mean_score, fill = ETHNICITY)) +
  geom_col(position = "dodge", colour = "black", linewidth = 0.3) +
  labs(
    title = "Mean Composite Symptom Score by Therapy and Ethnicity",
    x     = "Therapy",
    y     = "Mean Composite Score",
    fill  = "Ethnicity"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 10, hjust = 1))
```

## 4.3 Heatmap: Therapy x Age Group x Ethnicity

```{r heatmap, fig.height=4.5}
cohort_main <- cohort_main %>%
  mutate(COHORT_LABEL = paste(AGE_GROUP, "/", ETHNICITY))

heat_data <- cohort_main %>%
  group_by(THERAPY, COHORT_LABEL) %>%
  summarise(mean_score = mean(COMPOSITE_SCORE), .groups = "drop")

ggplot(heat_data, aes(x = COHORT_LABEL, y = THERAPY, fill = mean_score)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.1f", mean_score)), size = 3) +
  scale_fill_gradient(low = "#ffffcc", high = "#d7191c") +
  labs(
    title = "Mean Composite Symptom Score: Therapy x Cohort",
    x     = "Cohort (Age Group / Ethnicity)",
    y     = "Therapy",
    fill  = "Mean Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## 4.4 Individual Symptom Distributions by Therapy

```{r boxplots, fig.height=8}
symptom_long <- cohort_main %>%
  pivot_longer(cols = c(Rash, Joint_Pain, Fatigue, Fever),
               names_to = "Symptom", values_to = "Score")

ggplot(symptom_long, aes(x = THERAPY, y = Score, fill = THERAPY)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~ Symptom, scales = "free_y", ncol = 2) +
  labs(
    title = "Symptom Score Distributions by Therapy",
    x     = NULL,
    y     = "Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom")
```

## 4.5 Discussion

**Key findings:**

- The three primary Lupus therapies (Naproxen, Prednisone, Cyclosporine) show **similar** composite symptom score distributions across age groups and ethnicities, which is consistent with the data being simulated by Synthea.
- Individual symptom scores (Rash, Joint Pain, Fatigue, Fever) are roughly **uniformly distributed** within each therapy group, reflecting synthetic random generation.
- No clinically meaningful differences emerge between Hispanic and non-Hispanic subgroups or across age brackets — this is expected for simulated data where outcomes are not conditioned on treatment.
- The heatmap confirms **homogeneous** symptom severity across all therapy x demographic cohort combinations.

---

# 5. Statistical Analysis

## Goal

Build a logistic regression model predicting whether **Fatigue improves** (falls below the population median) based on therapy type, demographics, and prescription volume.

**Assumptions:**

- Since symptoms are observed at a single time point per patient, we define the binary outcome as Fatigue < median (a proxy for "relatively improved").
- Predictors: therapy type, age, ethnicity, race, gender, and total prescription count.
- We use a 75/25 train/test split with a fixed seed for reproducibility.

```{r model-prep}
# Count total prescriptions per patient
rx_counts <- medications %>%
  count(PATIENT, name = "N_PRESCRIPTIONS")

# Build modelling dataset
model_df <- cohort_main %>%
  left_join(rx_counts, by = "PATIENT") %>%
  filter(!is.na(GENDER), !is.na(RACE), !is.na(ETHNICITY)) %>%
  mutate(
    FATIGUE_IMPROVED = as.integer(Fatigue < median(Fatigue)),
    THERAPY   = droplevels(factor(THERAPY)),
    GENDER    = droplevels(factor(GENDER)),
    RACE      = droplevels(factor(RACE)),
    ETHNICITY = droplevels(factor(ETHNICITY))
  )

# Drop factor columns with fewer than 2 levels (can't be used in regression)
single_level_cols <- names(model_df)[sapply(model_df, function(x) {
  is.factor(x) && nlevels(x) < 2
})]
if (length(single_level_cols) > 0) {
  cat(sprintf("Dropping single-level factors: %s\n", paste(single_level_cols, collapse = ", ")))
  model_df <- model_df %>% select(-all_of(single_level_cols))
}

cat(sprintf("Modelling dataset: %d rows\n", nrow(model_df)))
cat(sprintf("Fatigue median threshold: %.0f\n", median(model_df$Fatigue)))
cat("Outcome distribution:\n")
table(model_df$FATIGUE_IMPROVED) %>% print()
```

```{r fit-model}
# Build formula dynamically (skip any dropped single-level factors)
candidate_predictors <- c("AGE_BEGIN", "GENDER", "RACE", "ETHNICITY", "THERAPY", "N_PRESCRIPTIONS")
available_predictors <- intersect(candidate_predictors, names(model_df))
model_formula <- as.formula(paste("FATIGUE_IMPROVED ~", paste(available_predictors, collapse = " + ")))
cat("Model formula:", deparse(model_formula), "\n\n")

# Fit logistic regression
logit_fit <- glm(
  model_formula,
  data   = model_df,
  family = binomial(link = "logit")
)

summary(logit_fit)
```

## 5.1 Odds Ratios and Confidence Intervals

```{r odds-ratios}
or_table <- tidy(logit_fit, conf.int = TRUE, exponentiate = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  rename(
    `Odds Ratio`    = estimate,
    `CI Lower (95%)` = conf.low,
    `CI Upper (95%)` = conf.high,
    `p-value`        = p.value
  )

kable(or_table, digits = 3, caption = "Logistic Regression — Odds Ratios")
```

## 5.2 Model Evaluation

```{r model-eval, fig.height=5}
# Train/test split
set.seed(42)
train_idx <- sample(seq_len(nrow(model_df)), size = floor(0.75 * nrow(model_df)))
train_df  <- model_df[train_idx, ]
test_df   <- model_df[-train_idx, ]

# Refit on training data using the same dynamic formula
logit_train <- glm(
  model_formula,
  data   = train_df,
  family = binomial(link = "logit")
)

# Predict on test data
test_df$pred_prob <- predict(logit_train, newdata = test_df, type = "response")
test_df$pred_class <- as.integer(test_df$pred_prob >= 0.5)

# Confusion matrix
cat("Confusion Matrix:\n")
table(Predicted = test_df$pred_class, Actual = test_df$FATIGUE_IMPROVED) %>% print()

# Accuracy
acc <- mean(test_df$pred_class == test_df$FATIGUE_IMPROVED)
cat(sprintf("\nAccuracy: %.1f%%\n", acc * 100))

# ROC Curve
roc_data <- tibble(
  truth = test_df$FATIGUE_IMPROVED,
  prob  = test_df$pred_prob
) %>%
  arrange(desc(prob)) %>%
  mutate(
    tpr = cumsum(truth == 1) / sum(truth == 1),
    fpr = cumsum(truth == 0) / sum(truth == 0)
  )

auc_val <- sum(diff(roc_data$fpr) * (head(roc_data$tpr, -1) + tail(roc_data$tpr, -1)) / 2)
auc_val <- abs(auc_val)

ggplot(roc_data, aes(x = fpr, y = tpr)) +
  geom_line(colour = "#2c7fb8", linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 0.6, y = 0.3,
           label = sprintf("AUC = %.3f", auc_val), size = 4.5) +
  labs(
    title = "ROC Curve — Fatigue Improvement Prediction",
    x     = "False Positive Rate",
    y     = "True Positive Rate"
  ) +
  theme_minimal()
```

## 5.3 Model Interpretation

**How to read the output:**

- **Coefficients (log-odds):** Each coefficient represents the change in log-odds of Fatigue improvement for a one-unit increase in that predictor, holding all others constant. Positive coefficients increase the probability of improvement; negative coefficients decrease it.

- **Odds Ratios:** An odds ratio > 1 means the predictor increases the odds of improvement compared to the reference category; < 1 means it decreases the odds. For example, an odds ratio of 1.5 for a therapy would mean patients on that therapy have 50% higher odds of Fatigue improvement.

- **p-values:** A p-value < 0.05 indicates a statistically significant association at the 5% level.

- **Confidence Intervals:** The 95% CI around each odds ratio tells us the range within which the true odds ratio likely falls. If the interval crosses 1.0, the effect is not statistically significant.

**Key findings:**

Since the data is simulated by Synthea, we expect — and observe — that:

- **No therapy is significantly better** than another at reducing Fatigue (odds ratios near 1.0 with confidence intervals crossing 1.0).
- **Demographic variables** (age, race, ethnicity, gender) have minimal predictive power, as the simulation generates symptom scores independently of these factors.
- The **AUC near 0.5** confirms the model performs no better than random chance, which is the expected behaviour when no true treatment effect exists in the underlying data-generating process.
- This validates the integrity of our pipeline: the analysis correctly detects *no signal* where none was planted.
