---
title: "Century Health ŌĆö Real World Evidence: Lupus Patient Analysis"
author: "Data Analyst Take-Home"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

# 0. Setup & Data Loading

```{r load-libraries}
library(tidyverse)
library(lubridate)
library(readxl)
library(scales)
library(patchwork)
library(broom)
library(testthat)
library(knitr)

# Load modular helper functions
source("R/helpers.R")

cat("All libraries and helpers loaded.\n")
```

```{r load-data}
raw <- load_all_datasets("data")

cat(sprintf("Patients:    %d x %d\n", nrow(raw$patients),    ncol(raw$patients)))
cat(sprintf("Encounters:  %d x %d\n", nrow(raw$encounters),  ncol(raw$encounters)))
cat(sprintf("Symptoms:    %d x %d\n", nrow(raw$symptoms),    ncol(raw$symptoms)))
cat(sprintf("Medications: %d x %d\n", nrow(raw$medications), ncol(raw$medications)))
cat(sprintf("Conditions:  %d x %d\n", nrow(raw$conditions),  ncol(raw$conditions)))
```

---

# 1. Data Assessment

## 1.1 Entity-Relationship Diagram

The five datasets are connected through **patient UUIDs** and **encounter UUIDs**:

```
                          ŌöīŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöÉ
                          Ōöé   PATIENTS   Ōöé
                          Ōöé (PATIENT_ID) Ōöé
                          ŌööŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌö¼ŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöś
                                 Ōöé  1
              ŌöīŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌö╝ŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöÉ
              Ōöé                  Ōöé                  Ōöé
              Ō¢╝ *                Ō¢╝ *                Ō¢╝ *
     ŌöīŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöÉ  ŌöīŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöÉ  ŌöīŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöÉ
     Ōöé  ENCOUNTERS    Ōöé  Ōöé  SYMPTOMS    Ōöé  Ōöé  MEDICATIONS Ōöé
     Ōöé  (Id)          Ōöé  Ōöé  (PATIENT)   Ōöé  Ōöé  (PATIENT,   Ōöé
     Ōöé  FK: PATIENT   Ōöé  Ōöé              Ōöé  Ōöé   ENCOUNTER) Ōöé
     ŌööŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌö¼ŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöś  ŌööŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöś  ŌööŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌö¼ŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöś
             Ōöé 1                                  Ōöé
             Ōöé            ŌöīŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöÉ        Ōöé
             ŌööŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌ¢║Ōöé  CONDITIONS  ŌöéŌŚäŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöś
                     *    Ōöé  (PATIENT,   Ōöé    *
                          Ōöé   ENCOUNTER) Ōöé
                          ŌööŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöĆŌöś
```

| From | To | Join Key |
|------|------|----------|
| Patients ŌåÆ Encounters | `PATIENT_ID` = `PATIENT` |
| Patients ŌåÆ Symptoms | `PATIENT_ID` = `PATIENT` |
| Patients ŌåÆ Medications | `PATIENT_ID` = `PATIENT` |
| Patients ŌåÆ Conditions | `PATIENT_ID` = `PATIENT` |
| Encounters ŌåÆ Medications | `Id` = `ENCOUNTER` |
| Encounters ŌåÆ Conditions | `Id` = `ENCOUNTER` |

> **Note:** The source files use a mix of CSV, Parquet, and Excel formats. The `load_all_datasets()` helper handles this transparently, converting Parquet to CSV on the fly if the `arrow` R package is unavailable.

## 1.2 Key Data Quality Issues Identified

| # | Issue | Impact | Resolution |
|---|-------|--------|------------|
| 1 | Mixed-case UUID keys across files | Joins fail silently | Normalise all IDs to lowercase |
| 2 | Same medication in 3+ casing variants | Inflated distinct-med counts | Standardise to uppercase |
| 3 | GENDER column entirely NA in symptoms | Missing demographic info | Backfill from patients table |
| 4 | Date columns stored as strings | Can't do time-based analysis | Parse to Date / POSIXct |
| 5 | SYMPTOMS is a packed `key:value` string | Not analysable directly | Parse into 4 numeric columns |
| 6 | Symptoms table has duplicate rows per patient | Overstates sample size | Filter to Lupus pathology, deduplicate |
| 7 | File format heterogeneity (CSV, Parquet, XLSX) | Loading complexity | Handled in `load_all_datasets()` |

---

# 2. Data Cleaning

```{r clean-data}
# Step 1: Normalise IDs to lowercase
datasets <- normalise_ids(raw)

# Step 2: Standardise medication names
datasets$medications <- standardise_medication_names(datasets$medications)
cat(sprintf("Medication descriptions: %d unique (after normalisation)\n",
            n_distinct(datasets$medications$DESCRIPTION)))

# Step 3: Parse dates (including DEATHDATE and conditions$STOP)
datasets <- parse_dates(datasets)

# Step 4: Parse symptom scores
datasets$symptoms <- parse_symptom_scores(datasets$symptoms)

# Step 5: Filter to Lupus-only rows and deduplicate to one row per patient
cat(sprintf("Symptoms before Lupus filter: %d rows, %d unique patients\n",
            nrow(datasets$symptoms), n_distinct(datasets$symptoms$PATIENT)))
datasets$symptoms <- filter_lupus_symptoms(datasets$symptoms)
cat(sprintf("Symptoms after  Lupus filter: %d rows, %d unique patients\n",
            nrow(datasets$symptoms), n_distinct(datasets$symptoms$PATIENT)))

# Step 6: Fill missing GENDER in symptoms from patients table
datasets$symptoms <- fill_symptom_gender(datasets$symptoms, datasets$patients)
cat(sprintf("Symptom GENDER nulls remaining: %d\n",
            sum(is.na(datasets$symptoms$GENDER))))

# Show parsed symptom scores
datasets$symptoms %>%
  select(PATIENT, PATHOLOGY, Rash, Joint_Pain, Fatigue, Fever) %>%
  head(5) %>%
  kable(caption = "Parsed Lupus symptom scores (first 5 rows)")
```

## 2.1 Data Integrity Unit Tests

```{r unit-tests, results='asis'}
run_integrity_tests(datasets)
cat("\n**All data integrity tests passed.**\n")
```

```{r extract-for-convenience}
# Extract cleaned tables for downstream analysis
patients    <- datasets$patients
encounters  <- datasets$encounters
symptoms    <- datasets$symptoms
medications <- datasets$medications
conditions  <- datasets$conditions

# Define the Lupus cohort patient IDs (patients with at least one Lupus condition)
lupus_patient_ids <- unique(conditions %>%
  filter(str_to_upper(DESCRIPTION) == "LUPUS ERYTHEMATOSUS") %>%
  pull(PATIENT))

lupus_patients <- patients %>% filter(PATIENT_ID %in% lupus_patient_ids)

cat(sprintf("Total patients in file: %d\n", n_distinct(patients$PATIENT_ID)))
cat(sprintf("Lupus cohort patients:  %d\n", n_distinct(lupus_patients$PATIENT_ID)))
```

---

# 3. Descriptive Analysis

All descriptive statistics below are scoped to the **Lupus cohort** ŌĆö patients with at least one Lupus erythematosus diagnosis in the conditions table.

## 3.1 Distinct Lupus Patients

```{r distinct-patients}
n_lupus <- n_distinct(lupus_patients$PATIENT_ID)
cat(sprintf("Distinct Lupus patients: %s\n", comma(n_lupus)))
```

## 3.2 Distinct Medications Over Time

```{r meds-over-time, fig.height=4}
# Restrict to medications for Lupus patients
lupus_meds <- medications %>% filter(PATIENT %in% lupus_patient_ids)

meds_over_time <- lupus_meds %>%
  mutate(YEAR_MONTH = floor_date(START, "month")) %>%
  group_by(YEAR_MONTH) %>%
  summarise(distinct_meds = n_distinct(DESCRIPTION), .groups = "drop")

ggplot(meds_over_time, aes(x = YEAR_MONTH, y = distinct_meds)) +
  geom_line(linewidth = 0.7, colour = "#2c7fb8") +
  geom_point(size = 1.5, colour = "#2c7fb8") +
  scale_y_continuous(breaks = scales::pretty_breaks()) +
  labs(
    title = "Distinct Medications Prescribed per Month (Lupus Cohort)",
    x     = "Date",
    y     = "Distinct Medications"
  ) +
  theme_minimal()
```

## 3.3 Lupus Cohort Distribution by Race and Gender

```{r pie-charts, fig.height=5}
# Race pie ŌĆö Lupus cohort only
race_data <- lupus_patients %>%
  count(RACE) %>%
  mutate(pct = n / sum(n),
         label = sprintf("%s\n%.1f%%", RACE, pct * 100))

p_race <- ggplot(race_data, aes(x = "", y = n, fill = RACE)) +
  geom_col(width = 1, colour = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5), size = 3) +
  labs(title = "Lupus Patients by Race", fill = "Race") +
  theme_void()

# Gender pie ŌĆö Lupus cohort only
gender_data <- lupus_patients %>%
  count(GENDER) %>%
  mutate(pct = n / sum(n),
         label = sprintf("%s\n%.1f%%", GENDER, pct * 100))

p_gender <- ggplot(gender_data, aes(x = "", y = n, fill = GENDER)) +
  geom_col(width = 1, colour = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5), size = 3.5) +
  labs(title = "Lupus Patients by Gender", fill = "Gender") +
  theme_void()

p_race + p_gender
```

## 3.4 Patients with All Four Symptom Categories >= 30

```{r high-symptoms}
# symptoms is already one row per Lupus patient after dedup
all_high <- symptoms %>%
  filter(Rash >= 30, Joint_Pain >= 30, Fatigue >= 30, Fever >= 30)

pct_high <- nrow(all_high) / nrow(symptoms) * 100

cat(sprintf(
  "Lupus patients with ALL four symptom scores >= 30: %d / %d (%.1f%%)\n",
  nrow(all_high), nrow(symptoms), pct_high
))
```

---

# 4. Cohort Analysis ŌĆö Treatments, Demographics & Outcomes

## Analysis Plan

We examine how **medication type**, **ethnicity**, and **age group** relate to symptom severity within the Lupus cohort.

**Cohort definitions:**

| Dimension | Groups |
|-----------|--------|
| Medication | Naproxen (NSAID), Prednisone (Corticosteroid), Cyclosporine (Immunosuppressant) |
| Ethnicity | Hispanic, Non-Hispanic |
| Age Group | Young (<30), Middle (30ŌĆō54), Older (55+) |

**Outcome measure:** Mean composite symptom score (mean of Rash, Joint Pain, Fatigue, Fever).

**Approach:**

1. Classify each medication into a therapeutic group.
2. Identify each patient's primary therapy (most frequently prescribed; ties broken deterministically by `slice_max`).
3. Assign age groups and compute composite symptom scores.
4. Compare outcomes across therapy ├Ś ethnicity ├Ś age group.

```{r build-cohort}
# Classify therapies (Lupus meds only)
lupus_meds <- lupus_meds %>%
  mutate(THERAPY = classify_therapy(DESCRIPTION))

cat("Therapy distribution (Lupus cohort prescriptions):\n")
lupus_meds %>% count(THERAPY, sort = TRUE) %>% kable()
```

```{r build-analysis-cohort}
# Each patient's primary therapy ŌĆö deterministic tie-breaking with slice_max
patient_therapy <- lupus_meds %>%
  count(PATIENT, THERAPY, name = "n_prescriptions") %>%
  group_by(PATIENT) %>%
  slice_max(n_prescriptions, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(PATIENT, THERAPY)

# Add composite score and age group to the deduplicated symptoms table
symptoms <- symptoms %>%
  add_composite_score() %>%
  mutate(AGE_GROUP = assign_age_group(AGE_BEGIN))

# Join symptoms with primary therapy
cohort <- symptoms %>%
  inner_join(patient_therapy, by = "PATIENT")

# Focus on the three main Lupus therapies
main_therapies <- c("Naproxen (NSAID)", "Prednisone (Corticosteroid)",
                    "Cyclosporine (Immunosuppressant)")
cohort_main <- cohort %>% filter(THERAPY %in% main_therapies)

cat(sprintf("Analysis cohort: %d patients (one row each)\n",
            nrow(cohort_main)))
cat(sprintf("  Naproxen:      %d\n", sum(cohort_main$THERAPY == main_therapies[1])))
cat(sprintf("  Prednisone:    %d\n", sum(cohort_main$THERAPY == main_therapies[2])))
cat(sprintf("  Cyclosporine:  %d\n", sum(cohort_main$THERAPY == main_therapies[3])))
```

## 4.1 Symptom Severity by Therapy and Age Group

```{r therapy-age-bar, fig.height=5}
summary_age <- cohort_main %>%
  group_by(THERAPY, AGE_GROUP) %>%
  summarise(mean_score = mean(COMPOSITE_SCORE), .groups = "drop")

ggplot(summary_age, aes(x = THERAPY, y = mean_score, fill = AGE_GROUP)) +
  geom_col(position = "dodge", colour = "black", linewidth = 0.3) +
  labs(
    title = "Mean Composite Symptom Score by Therapy and Age Group",
    x     = "Therapy",
    y     = "Mean Composite Score",
    fill  = "Age Group"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 10, hjust = 1))
```

## 4.2 Symptom Severity by Therapy and Ethnicity

```{r therapy-ethnicity-bar, fig.height=5}
summary_eth <- cohort_main %>%
  group_by(THERAPY, ETHNICITY) %>%
  summarise(mean_score = mean(COMPOSITE_SCORE), .groups = "drop")

ggplot(summary_eth, aes(x = THERAPY, y = mean_score, fill = ETHNICITY)) +
  geom_col(position = "dodge", colour = "black", linewidth = 0.3) +
  labs(
    title = "Mean Composite Symptom Score by Therapy and Ethnicity",
    x     = "Therapy",
    y     = "Mean Composite Score",
    fill  = "Ethnicity"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 10, hjust = 1))
```

## 4.3 Heatmap: Therapy ├Ś Age Group ├Ś Ethnicity

```{r heatmap, fig.height=4.5}
cohort_main <- cohort_main %>%
  mutate(COHORT_LABEL = paste(AGE_GROUP, "/", ETHNICITY))

heat_data <- cohort_main %>%
  group_by(THERAPY, COHORT_LABEL) %>%
  summarise(mean_score = mean(COMPOSITE_SCORE), .groups = "drop")

ggplot(heat_data, aes(x = COHORT_LABEL, y = THERAPY, fill = mean_score)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.1f", mean_score)), size = 3) +
  scale_fill_gradient(low = "#ffffcc", high = "#d7191c") +
  labs(
    title = "Mean Composite Symptom Score: Therapy ├Ś Cohort",
    x     = "Cohort (Age Group / Ethnicity)",
    y     = "Therapy",
    fill  = "Mean Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## 4.4 Individual Symptom Distributions by Therapy

```{r boxplots, fig.height=8}
symptom_long <- cohort_main %>%
  pivot_longer(cols = c(Rash, Joint_Pain, Fatigue, Fever),
               names_to = "Symptom", values_to = "Score")

ggplot(symptom_long, aes(x = THERAPY, y = Score, fill = THERAPY)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~ Symptom, scales = "free_y", ncol = 2) +
  labs(
    title = "Symptom Score Distributions by Therapy",
    x     = NULL,
    y     = "Score"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom")
```

## 4.5 Discussion

```{r cohort-summary-stats}
# Summarise actual values so the narrative is grounded in outputs, not assumptions
overall_means <- cohort_main %>%
  group_by(THERAPY) %>%
  summarise(
    n = n(),
    mean_composite = round(mean(COMPOSITE_SCORE), 1),
    sd_composite   = round(sd(COMPOSITE_SCORE), 1),
    .groups = "drop"
  )

kable(overall_means,
      col.names = c("Therapy", "N", "Mean Composite", "SD"),
      caption   = "Summary statistics by therapy")

# Check whether age-group or ethnicity differences are large
age_spread <- summary_age %>%
  group_by(THERAPY) %>%
  summarise(range = round(max(mean_score) - min(mean_score), 1), .groups = "drop")

eth_spread <- summary_eth %>%
  group_by(THERAPY) %>%
  summarise(range = round(max(mean_score) - min(mean_score), 1), .groups = "drop")

cat("\nMax within-therapy difference across age groups:\n")
print(age_spread)
cat("\nMax within-therapy difference across ethnicities:\n")
print(eth_spread)
```

**Observations based on the summaries above:**

- Composite symptom scores are similar across therapies (see the table above for the actual means and standard deviations). If the per-therapy means and SDs are nearly identical, this indicates no observable treatment differentiation in symptom severity.
- The within-therapy spread across age groups and ethnicities is reported above. Small ranges (a few points) indicate demographic subgroups do not meaningfully differ within a therapy.
- The box plots show the full distributions. If the medians and interquartile ranges overlap substantially, this further supports the absence of a treatment-specific signal.
- These patterns are consistent with Synthea's simulation approach, which generates symptom scores independently of treatment assignment.

---

# 5. Statistical Analysis

## Goal

Build a logistic regression model to assess whether therapy type and demographics are associated with **lower Fatigue scores** in the Lupus cohort.

**Important framing:** Because symptoms are observed at a single cross-sectional time point (not longitudinally), this model does **not** estimate treatment effect or improvement over time. Instead, it tests whether having low Fatigue (below the cohort median) is associated with therapy type, demographics, or prescription volume.

**Predictors:** therapy type, age, ethnicity, race, gender, total prescription count.

**Train/test split:** 75/25, stratified at the **patient level** (each patient appears in exactly one set since we have one row per patient after deduplication).

```{r model-prep}
# Count total prescriptions per patient
rx_counts <- lupus_meds %>%
  count(PATIENT, name = "N_PRESCRIPTIONS")

# Build modelling dataset
model_df <- cohort_main %>%
  left_join(rx_counts, by = "PATIENT") %>%
  filter(!is.na(GENDER), !is.na(RACE), !is.na(ETHNICITY)) %>%
  mutate(
    LOW_FATIGUE = as.integer(Fatigue < median(Fatigue)),
    THERAPY     = droplevels(factor(THERAPY)),
    GENDER      = droplevels(factor(GENDER)),
    RACE        = droplevels(factor(RACE)),
    ETHNICITY   = droplevels(factor(ETHNICITY))
  )

# Drop factor columns with fewer than 2 levels (can't be used in regression)
single_level_cols <- names(model_df)[sapply(model_df, function(x) {
  is.factor(x) && nlevels(x) < 2
})]
if (length(single_level_cols) > 0) {
  cat(sprintf("Dropping single-level factors: %s\n",
              paste(single_level_cols, collapse = ", ")))
  model_df <- model_df %>% select(-all_of(single_level_cols))
}

cat(sprintf("Modelling dataset: %d patients\n", nrow(model_df)))
cat(sprintf("Fatigue median threshold: %.0f\n", median(model_df$Fatigue)))
cat("Outcome distribution (LOW_FATIGUE):\n")
table(model_df$LOW_FATIGUE) %>% print()
```

```{r fit-model}
# Build formula dynamically (skip any dropped single-level factors)
candidate_predictors <- c("AGE_BEGIN", "GENDER", "RACE", "ETHNICITY",
                          "THERAPY", "N_PRESCRIPTIONS")
available_predictors <- intersect(candidate_predictors, names(model_df))
model_formula <- as.formula(
  paste("LOW_FATIGUE ~", paste(available_predictors, collapse = " + "))
)
cat("Model formula:", deparse(model_formula), "\n\n")

# Fit logistic regression on full dataset for coefficient inspection
logit_fit <- glm(model_formula, data = model_df, family = binomial(link = "logit"))

summary(logit_fit)
```

## 5.1 Odds Ratios and Confidence Intervals

```{r odds-ratios}
or_table <- tidy(logit_fit, conf.int = TRUE, exponentiate = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  rename(
    `Odds Ratio`     = estimate,
    `CI Lower (95%)` = conf.low,
    `CI Upper (95%)` = conf.high,
    `p-value`        = p.value
  )

kable(or_table, digits = 3, caption = "Logistic Regression ŌĆö Odds Ratios")
```

## 5.2 Model Evaluation

```{r model-eval, fig.height=5}
# Patient-level train/test split (one row per patient, so row = patient)
set.seed(42)
train_idx <- sample(seq_len(nrow(model_df)), size = floor(0.75 * nrow(model_df)))
train_df  <- model_df[train_idx, ]
test_df   <- model_df[-train_idx, ]

cat(sprintf("Train: %d patients, Test: %d patients\n",
            nrow(train_df), nrow(test_df)))

# Refit on training data
logit_train <- glm(model_formula, data = train_df, family = binomial(link = "logit"))

# Predict on test data
test_df$pred_prob  <- predict(logit_train, newdata = test_df, type = "response")
test_df$pred_class <- as.integer(test_df$pred_prob >= 0.5)

# Confusion matrix
cat("\nConfusion Matrix:\n")
table(Predicted = test_df$pred_class, Actual = test_df$LOW_FATIGUE) %>% print()

# Accuracy
acc <- mean(test_df$pred_class == test_df$LOW_FATIGUE)
cat(sprintf("\nAccuracy: %.1f%%\n", acc * 100))

# ROC Curve
roc_data <- tibble(
  truth = test_df$LOW_FATIGUE,
  prob  = test_df$pred_prob
) %>%
  arrange(desc(prob)) %>%
  mutate(
    tpr = cumsum(truth == 1) / sum(truth == 1),
    fpr = cumsum(truth == 0) / sum(truth == 0)
  )

auc_val <- abs(sum(diff(roc_data$fpr) *
  (head(roc_data$tpr, -1) + tail(roc_data$tpr, -1)) / 2))

ggplot(roc_data, aes(x = fpr, y = tpr)) +
  geom_line(colour = "#2c7fb8", linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 0.6, y = 0.3,
           label = sprintf("AUC = %.3f", auc_val), size = 4.5) +
  labs(
    title = "ROC Curve ŌĆö Low Fatigue Classification",
    x     = "False Positive Rate",
    y     = "True Positive Rate"
  ) +
  theme_minimal()
```

## 5.3 Model Interpretation

**How to read the output:**

- **Coefficients (log-odds):** Each coefficient represents the change in log-odds of having low Fatigue for a one-unit increase in that predictor, holding all others constant. Positive values increase the probability; negative values decrease it.

- **Odds Ratios:** An OR > 1 means the predictor increases the odds of low Fatigue relative to the reference category; OR < 1 means it decreases the odds.

- **p-values:** Values below 0.05 indicate statistical significance at the 5% level.

- **Confidence Intervals:** If the 95% CI for an odds ratio crosses 1.0, the association is not statistically significant.

```{r interpret-results, results='asis'}
# Dynamically report findings based on actual model output
n_sig <- sum(or_table$`p-value` < 0.05, na.rm = TRUE) - 1L  # exclude intercept

cat("**Key findings from this model:**\n\n")

if (n_sig == 0) {
  cat("- No predictor reaches statistical significance at the 5% level.\n")
  cat("- This means therapy type, demographics, and prescription volume are ",
      "not meaningfully associated with Fatigue level in this cohort.\n")
} else {
  sig_terms <- or_table %>% filter(`p-value` < 0.05, term != "(Intercept)")
  cat(sprintf("- %d predictor(s) reached significance at p < 0.05:\n", n_sig))
  for (i in seq_len(nrow(sig_terms))) {
    cat(sprintf("  - **%s**: OR = %.2f (95%% CI: %.2fŌĆō%.2f, p = %.3f)\n",
                sig_terms$term[i], sig_terms$`Odds Ratio`[i],
                sig_terms$`CI Lower (95%)`[i], sig_terms$`CI Upper (95%)`[i],
                sig_terms$`p-value`[i]))
  }
}

cat(sprintf("\n- **AUC = %.3f** ŌĆö ", auc_val))
if (auc_val < 0.55) {
  cat("near 0.5, indicating the model has no meaningful discriminatory power.\n")
} else if (auc_val < 0.7) {
  cat("indicating weak discriminatory power.\n")
} else {
  cat("indicating moderate to good discriminatory power.\n")
}

cat(sprintf("- **Accuracy = %.1f%%** on the held-out test set.\n\n", acc * 100))

cat("**Important caveats:**\n\n")
cat("- This is a **cross-sectional** analysis, not longitudinal. ",
    "The outcome (low vs. high Fatigue) does not represent change over time.\n")
cat("- Patients were not randomly assigned to therapies, so confounding is possible.\n")
cat("- The data is simulated by Synthea; any patterns (or lack thereof) reflect ",
    "the simulation's data-generating process rather than real clinical relationships.\n")
```
